Weakly Nonlinear MRI: Notes

### 2014-11-15 notes/mri_notes.tex
x1 -> x0
hMRI matrix form
tuning and multiscale hmri
hmri perturbation series expansion
more equations
hmri additions if the azimuthal term is constant in cartesian coordinates

\begin{equation}
\mathbf{B} = B_0 \hat{z} + \xi B_0 \left(\frac{x_1}{x}\right) \hat{y} + \mathbf{B}_1
\end{equation}

@@

\beq
\left(\nabla \times \left(B_0 \hat{z} + \xi B_0 \left(\frac{x_1}{x}\right) \hat{y} + \mathbf{B}_1
\right) \right) \times \left(B_0 \hat{z} + \xi B_0 \left(\frac{x_1}{x}\right) \hat{y} + \mathbf{B}_1
\right)
\eeq

@@

\beq
\left(\nabla \times B_0 \hat{z} \right) \times \xi B_0 \left(\frac{x_1}{x}\right)\hat{y} = 0
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \left(\frac{x_1}{x}\right) \hat{y} \right) \times B_0 \hat{z} = \left(\partial_x \left(\xi B_0 \left(\frac{x_1}{x}\right)\right)\hat{z}\right) \times B_0 \hat{z} = 0
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \frac{x_1}{x} \hat{y} \right) \times \left( \xi B_0 \frac{x_1}{x}\hat{y}\right) = \xi^2 B_0^2 \frac{x_1^2}{x^3} \hat{x}
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \frac{x_1}{x} \hat{y}\right) \times \mathbf{B_1} = -\xi B_0 \frac{x_1}{x^2} \left(B_{1x} \hat{y} - B_{1y} \hat{x}\right)
\eeq

@@

\beq
\left(\nabla \times \mathbf{B}_1 \right) \times \left(\xi B_0 \frac{x_1}{x} \hat{y}\right) = - \partial_z B_{1y} \xi B_0 \frac{x_1}{x} \hat{z} - \partial_x B_{1y} \xi B_0 \frac{x_1}{x} \hat{x}
\eeq

@@

\beq
\label{momeq_added}
\xi^2 B_0^2 \frac{x_1^2}{x^3} \hat{x} \, - \,  \xi B_0 \frac{x_1}{x^2} B_{1x} \hat{y} \, + \, \xi B_0 \frac{x_1}{x^2} B_{1y} \hat{x} \, - \, \partial_z B_{1y} \xi B_0 \frac{x_1}{x} \hat{z} \, - \, \partial_x B_{1y} \xi B_0 \frac{x_1}{x} \hat{x}
\eeq

@@

\beq
\left( \left( \mathbf{B_0} \hat{z} + \xi B_0 \frac{x_1}{x} \hat{y} + \mathbf{B_1} \right) \cdot \nabla \right) \left(-q \Omega_0 x \hat{y} + \mathbf{u_1}\right) = \xi B_0 \frac{x_1}{x} \partial_y \mathbf{u_1}
\eeq

@@

\beq
-\left( \left( -q \Omega_0 x \hat{y} + \mathbf{u_1}\right) \cdot \nabla\right) \left( B_0 \hat{z} + \xi B_0 \frac{x_1}{x} \hat{y} + \mathbf{B_1}\right)   \, = \, \xi B_0 u_{1x} \frac{x_1}{x^2} \hat{y}
\eeq

@@

\beq
\frac{1}{\reym} \xi B_0 \left( \frac{2 x_1}{x^3}\right) \hat{y}
\eeq

@@

\beq
\xi B_0 \frac{x_1}{x} \partial_y \mathbf{u_1} \, + \, \xi B_0 u_{1x} \frac{x_1}{x^2} \hat{y} \, + \, \frac{1}{\reym} \xi B_0 \left( \frac{2 x_1}{x^3}\right) \hat{y}
\eeq

@@

\beq
\label{induc_eq_added}
\xi B_0 \frac{x_1}{x^2} \partial_z \Psi \hat{y} + \frac{1}{\reym} \xi B_0 \frac{2 x_1}{x^3} \hat{y}
\eeq

@@

\beq
[B]: \xi B_0 \frac{x_1}{x^2} \partial_z \Psi + \frac{1}{\reym} \xi B_0 \frac{2 x_1}{x^3}
\eeq

@@

>>>>

\begin{equation}
\mathbf{B} = B_0 \hat{z} + \xi B_0 \left(\frac{x_0}{x}\right) \hat{y} + \mathbf{B}_1
\end{equation}

@@

\beq
\left(\nabla \times \left(B_0 \hat{z} + \xi B_0 \left(\frac{x_0}{x}\right) \hat{y} + \mathbf{B}_1
\right) \right) \times \left(B_0 \hat{z} + \xi B_0 \left(\frac{x_0}{x}\right) \hat{y} + \mathbf{B}_1
\right)
\eeq

@@

\beq
\left(\nabla \times B_0 \hat{z} \right) \times \xi B_0 \left(\frac{x_0}{x}\right)\hat{y} = 0
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \left(\frac{x_0}{x}\right) \hat{y} \right) \times B_0 \hat{z} = \left(\partial_x \left(\xi B_0 \left(\frac{x_0}{x}\right)\right)\hat{z}\right) \times B_0 \hat{z} = 0
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \frac{x_0}{x} \hat{y} \right) \times \left( \xi B_0 \frac{x_0}{x}\hat{y}\right) = \xi^2 B_0^2 \frac{x_0^2}{x^3} \hat{x}
\eeq

@@

\beq
\left(\nabla \times \xi B_0 \frac{x_0}{x} \hat{y}\right) \times \mathbf{B_1} = -\xi B_0 \frac{x_0}{x^2} \left(B_{1x} \hat{y} - B_{1y} \hat{x}\right)
\eeq

@@

\beq
\left(\nabla \times \mathbf{B}_1 \right) \times \left(\xi B_0 \frac{x_0}{x} \hat{y}\right) = - \partial_z B_{1y} \xi B_0 \frac{x_0}{x} \hat{z} - \partial_x B_{1y} \xi B_0 \frac{x_0}{x} \hat{x}
\eeq

@@

\beq
\label{momeq_added}
\xi^2 B_0^2 \frac{x_0^2}{x^3} \hat{x} \, - \,  \xi B_0 \frac{x_0}{x^2} B_{1x} \hat{y} \, + \, \xi B_0 \frac{x_0}{x^2} B_{1y} \hat{x} \, - \, \partial_z B_{1y} \xi B_0 \frac{x_0}{x} \hat{z} \, - \, \partial_x B_{1y} \xi B_0 \frac{x_0}{x} \hat{x}
\eeq

@@

\beq
\left( \left( \mathbf{B_0} \hat{z} + \xi B_0 \frac{x_0}{x} \hat{y} + \mathbf{B_1} \right) \cdot \nabla \right) \left(-q \Omega_0 x \hat{y} + \mathbf{u_1}\right) = \xi B_0 \frac{x_0}{x} \partial_y \mathbf{u_1}
\eeq

@@

\beq
-\left( \left( -q \Omega_0 x \hat{y} + \mathbf{u_1}\right) \cdot \nabla\right) \left( B_0 \hat{z} + \xi B_0 \frac{x_0}{x} \hat{y} + \mathbf{B_1}\right)   \, = \, \xi B_0 u_{1x} \frac{x_0}{x^2} \hat{y}
\eeq

@@

\beq
\frac{1}{\reym} \xi B_0 \left( \frac{2 x_0}{x^3}\right) \hat{y}
\eeq

@@

\beq
\xi B_0 \frac{x_0}{x} \partial_y \mathbf{u_1} \, + \, \xi B_0 u_{1x} \frac{x_0}{x^2} \hat{y} \, + \, \frac{1}{\reym} \xi B_0 \left( \frac{2 x_0}{x^3}\right) \hat{y}
\eeq

@@

\beq
\label{induc_eq_added}
\xi B_0 \frac{x_0}{x^2} \partial_z \Psi \hat{y} + \frac{1}{\reym} \xi B_0 \frac{2 x_0}{x^3} \hat{y}
\eeq

@@

\beq
[B]: \xi B_0 \frac{x_0}{x^2} \partial_z \Psi + \frac{1}{\reym} \xi B_0 \frac{2 x_0}{x^3}
\eeq

@@

We now rewrite these terms in matrix form. We will have a matrix for the variable terms and a constant vector.

\beq
\mathcal{H} = \left[ \begin{matrix}  0 & 0 & 0 & x_0\left(\frac{1}{x^2} - \frac{1}{x}\partial_x\right)\\
0 & 0 & -\frac{x_0}{x^2} & 0 \\ 
0 & 0 & 0 & 0 \\
\frac{x_0}{x^2} & 0 & 0 & 0 \\
\end{matrix} \right] \\
\eeq

and

\beq
\mathbf{C} = \left[\begin{matrix} 0 \\
0 \\
0 \\
\frac{1}{\reym} \frac{2 x_0}{x^3} \\
\end{matrix} \right] \\
\eeq

And our additional terms become:

\beq
\xi B_0 \mathcal{H} \partial_z \mathbf{V} + \xi B_0 \mathbf{C}
\eeq

Destabilizing substitution $B_0 = 1 - \epsilon^2$:

\beq
\xi \left(1 - \epsilon^2\right) \mathcal{H} \partial_z \mathbf{V} + \xi \left(1 - \epsilon^2\right) \mathbf{C}
\eeq

\beq
= \xi \mathcal{H} \partial_z \mathbf{V}  - \epsilon^2 \xi \mathcal{H} \partial_z \mathbf{V} + \xi \mathbf{C} - \xi \epsilon^2 \mathbf{C}
\eeq

Multiscale expansion ($\partial_z \rightarrow \partial_z + \epsilon \partial_Z$):

\beq
= \xi \mathcal{H} \partial_z \mathbf{V} + \epsilon \xi \mathcal{H} \partial_Z \mathbf{V} - \epsilon^2 \xi \mathcal{H} \partial_z \mathbf{V} - \epsilon^3 \xi \mathcal{H} \partial_Z \mathbf{V} + \xi \mathbf{C} - \xi \epsilon^2 \mathbf{C}
\eeq

And finally, the perturbation series expansion $\mathbf{V} = \epsilon \mathbf{V_1} + \epsilon^2 \mathbf{V_2} + \epsilon^3 \mathbf{V_3}$. I include terms up to $\epsilon^3$:

\beq
\epsilon \xi \mathcal{H} \partial_z \mathbf{V_1} + \epsilon^2 \xi \mathcal{H} \partial_z \mathbf{V_2} + \epsilon^3 \xi \mathcal{H} \partial_z \mathbf{V_3} + \epsilon^2 \xi \mathcal{H} \partial_Z \mathbf{V_1} + \epsilon^3 \xi \mathcal{H} \partial_Z \mathbf{V_2} - \epsilon^3 \xi \mathcal{H} \partial_z \mathbf{V_1} + \xi \mathbf{C} - \xi \epsilon^2 \mathbf{C}
\eeq

Separated by order:

$\mathcal{O}(0)$:

\beq
\xi \mathbf{C}
\eeq

$\mathcal{O}(\epsilon)$:

\beq
\epsilon \xi \mathcal{H} \partial_z \mathbf{V_1}
\eeq

$\mathcal{O}(\epsilon^2)$:

\beq
\epsilon^2 \xi \mathcal{H} \partial_z \mathbf{V_2} + \epsilon^2 \xi \mathcal{H} \partial_Z \mathbf{V_1} - \xi \epsilon^2 \mathbf{C}
\eeq

$\mathcal{O}(\epsilon^3)$:

\beq
\epsilon^3 \xi \mathcal{H} \partial_z \mathbf{V_3} + \epsilon^3 \xi \mathcal{H} \partial_Z \mathbf{V_2} - \epsilon^3 \xi \mathcal{H} \partial_z \mathbf{V_1}
\eeq

Meaning that the equations for the hMRI become: \\

$\mathcal{O}(\epsilon)$:

\beq
\mathcal{L}\mathbf{V_1} + \xi \mathcal{H} \partial_z \mathbf{V_1} = 0
\eeq

$\mathcal{O}(\epsilon^2)$:

\beq
\mathcal{L} \mathbf{V_2} = \mathbf{N_2} - \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V_1} + \xi \mathcal{H} \partial_z \mathbf{V_2} + \xi \mathcal{H} \partial_Z \mathbf{V_1} - \xi \mathbf{C}
\eeq

$\mathcal{O}(\epsilon^3)$:

\beq
\mathcal{D}\partial_T \mathbf{V_1} + \mathbf{N_3} = \mathcal{L} \mathbf{V_3} + \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V_2} + \widetilde{\mathcal{L}}_2 \partial_Z^2 \mathbf{V_1} - \partial_z \mathcal{X} \mathbf{V_1} - \partial_z^3 \mathcal{L}_3 \mathbf{V_1} +  \xi \mathcal{H} \partial_z \mathbf{V_3} + \xi \mathcal{H} \partial_Z \mathbf{V_2} -  \xi \mathcal{H} \partial_z \mathbf{V_1}
\eeq

If instead we Taylor expand the azimuthal field term to obtain a helical MRI background of:

\beq
\mathbf{B} = B_0 \hat{z} + \xi B_0 \hat{y} + \mathbf{B_1},
\eeq

we don't gain nearly as many terms with the hMRI. The momentum equation picks up only

\beq
-\xi B_0 \partial_z B_{1y} \hat{z} - \xi B_0 \partial_x B_{1y} \hat{x},
\eeq

and the induction equation picks up nothing if the perturbations are axisymmetric (the only term gained is $\xi B_0 \partial_y \mathbf{u_1}$, and $\partial_y \rightarrow 0$).

Taking $\partial_z$ of the [x] momentum equation and $\partial_x$ of the [z] momentum equation, the momentum equation hMRI addition becomes

\beq
-\xi B_0 \partial_x \partial_z B_{1y} \hat{z} - \xi B_0 \partial_x \partial_z B_{1y} \hat{x}.
\eeq
